---
title: "Data Exploration"
author: "Michael Green"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Load Packages and Data

```{r entry}
library(tidyverse)
library(tableone)
library(tidymodels)
# loading raw NAMCS data which I converted to csv
data_raw <- read.csv("Data/NAMCS2019.csv")

# loading pdf file of 20 pages from NAMCS data documentation that listed the key for the different medical codes
medcodes_raw <- read.csv("Data/NAMCS Drug Entry Codes.csv")

#Making the variables interpretable
data_clean <- data_raw %>%
  select(Age = AGE, 
         `Race and Ethnicity` = RACERETH, 
         `Smoking History` = EVERTOBAC, 
         `Major Reason for Visit` = MAJOR, 
         `Primary Reason for Visit` = RFV1, 
         `Secondary Reason for Visit` = RFV2, 
         `Primary Care` = PRIMCARE, 
         Hypertension = HTN, 
         `Chronic Kidney Disease` = CKD, 
         `Congestive Heart Failure` = CHF, 
         `Diabetes Type 1` = DIABTYP1, 
         `Diabetes Type 2` = DIABTYP2, 
         Obesity = OBESITY, 
         `Systolic Blood Pressure` = BPSYS, 
         Medication = MED, 
         `Number Medications` = NUMMED, 
         `Time with Doctor` = TIMEMD, 
         `Recommended for ED or ER` = ERADMHOS)

#Telling RStudio to treat these continuous variables as numeric
data_clean <- data_clean %>%
  mutate(Age = as.numeric(Age),
         `Number Medications` = as.numeric(`Number Medications`),
         `Time with Doctor` = as.numeric(`Time with Doctor`),
         `Systolic Blood Pressure` = as.numeric(`Systolic Blood Pressure`))

#Reducing the amount of category options for reason for visi (originally above 80), then dropping unknown and blank values for Smoking History, Primary Care, and Medication variables
data_clean <- data_clean %>%
  filter(`Smoking History` != "Unknown") %>%
  filter(!(`Primary Care` %in% c("Blank", "Unknown"))) %>%
  filter(`Medication` != "Entire item blank") %>%
  mutate(`Primary Reason for Visit` = if_else(`Primary Reason for Visit` %in% c("Hypertension", "Blood pressure test"), 
                                              `Primary Reason for Visit`, "other"),
         `Secondary Reason for Visit` = if_else(`Secondary Reason for Visit` %in% c("Hypertension", "Blood pressure test"), 
                                                `Secondary Reason for Visit`, "other"))

#Making the variables readable and removing excess space on observations
medcodes_clean <- medcodes_raw %>%
  select(Medication.ID, Medication.Name) %>%
  rename(`Rx ID Number` = Medication.ID, `Rx Name` = Medication.Name) %>%
  mutate(`Rx ID Number` = trimws(`Rx ID Number`),
         `Rx Name` = trimws(`Rx Name`))

#Telling RStudio to treat these continuous variables as numeric
data_clean <- data_clean %>%
  mutate(Age = as.numeric(Age),
         `Number Medications` = as.numeric(`Number Medications`),
         `Time with Doctor` = as.numeric(`Time with Doctor`),
         `Systolic Blood Pressure` = as.numeric(`Systolic Blood Pressure`))

#Reducing the amount of category options for reason for visi (originally above 80), dropping unknown and blank values for Smoking History, Primary Care, and Medication variables
data_clean <- data_clean %>%
  filter(`Smoking History` != "Unknown") %>%
  filter(!(`Primary Care` %in% c("Blank", "Unknown"))) %>%
  filter(`Medication` != "Entire item blank") %>%
  mutate(`Primary Reason for Visit` = if_else(`Primary Reason for Visit` %in% c("Hypertension", "Blood pressure test"), 
                                              `Primary Reason for Visit`, "other"),
         `Secondary Reason for Visit` = if_else(`Secondary Reason for Visit` %in% c("Hypertension", "Blood pressure test"), 
                                                `Secondary Reason for Visit`, "other"))

#Making the variables readbable and removing excess space on observations
medcodes_clean <- medcodes_raw %>%
  select(Medication.ID, Medication.Name) %>%
  rename(`Rx ID Number` = Medication.ID, `Rx Name` = Medication.Name) %>%
  mutate(`Rx ID Number` = trimws(`Rx ID Number`),
         `Rx Name` = trimws(`Rx Name`))

```


## Table 1 with 8250 Observations

```{r entry 3}


# Continuous and categorical variables
cont_vars <- c("Age", "Number Medications", "Time with Doctor", "Systolic Blood Pressure")
cat_vars <- c("Race and Ethnicity", "Smoking History", "Major Reason for Visit", "Primary Care",
              "Hypertension", "Chronic Kidney Disease", "Congestive Heart Failure", 
              "Diabetes Type 1", "Diabetes Type 2", "Obesity", "Medication", 
              "Recommended for ED or ER")

# Create the summary table
my_table <- CreateTableOne(vars = c(cont_vars, cat_vars), data = data_clean, factorVars = cat_vars)

# Print the table
print(my_table, smd = TRUE)

```

## Creating data sets based on non-surgery related visits, and people with primary or secondary reason of visit as Hypertension or Blood pressure test

```{r entry 4}

data_nonsurgery <- data_clean %>%
  filter(`Major Reason for Visit` %in% c("Chronic problem, routine",
                                         "New problem (less than 3 mos. onset)",
                                         "Chronic problem, flare-up",
                                         "Preventive care"))
data_htnvisits <- data_nonsurgery %>%
  filter(`Primary Reason for Visit` %in% c("Hypertension", "Blood pressure test") |
           `Secondary Reason for Visit` %in% c("Hypertension", "Blood pressure test"))

data_htnyes <- data_nonsurgery %>%
  filter(`Hypertension` %in% c("Yes")) 

```

```{r, include=FALSE}
# Create a vector with old and new variable names
old_med_names <- paste0("MED", 1:30)
new_med_names <- paste0("Medication #", 1:30)
old_ncmed_names <- paste0("NCMED", 1:30)
new_ncmed_names <- paste0("New or Continued Medication #", 1:30)

data_meds <- data_raw %>%
  select(all_of(c(old_med_names, old_ncmed_names)),
         Age = AGE, 
         `Race and Ethnicity` = RACERETH, 
         `Smoking History` = EVERTOBAC, 
         `Major Reason for Visit` = MAJOR, 
         `Primary Reason for Visit` = RFV1, 
         `Secondary Reason for Visit` = RFV2, 
         `Primary Care` = PRIMCARE, 
         Hypertension = HTN, 
         `Chronic Kidney Disease` = CKD, 
         `Congestive Heart Failure` = CHF, 
         `Diabetes Type 1` = DIABTYP1, 
         `Diabetes Type 2` = DIABTYP2, 
         Obesity = OBESITY, 
         `Systolic Blood Pressure` = BPSYS, 
         Medication = MED, 
         `Number Medications` = NUMMED, 
         `Time with Doctor` = TIMEMD, 
         `Recommended for ED or ER` = ERADMHOS) %>%
  rename_with(~ifelse(. %in% old_med_names, new_med_names[match(., old_med_names)], 
                       ifelse(. %in% old_ncmed_names, new_ncmed_names[match(., old_ncmed_names)], .))) %>%
  mutate(Age = as.numeric(Age),
         `Number Medications` = as.numeric(`Number Medications`),
         `Time with Doctor` = as.numeric(`Time with Doctor`),
         `Systolic Blood Pressure` = as.numeric(`Systolic Blood Pressure`)) %>%
  filter(`Smoking History` != "Unknown") %>%
  filter(!(`Primary Care` %in% c("Blank", "Unknown"))) %>%
  filter(`Medication` != "Entire item blank") %>%
  mutate(`Primary Reason for Visit` = if_else(`Primary Reason for Visit` %in% c("Hypertension", "Blood pressure test"), 
                                              `Primary Reason for Visit`, "other"),
         `Secondary Reason for Visit` = if_else(`Secondary Reason for Visit` %in% c("Hypertension", "Blood pressure test"), 
                                                `Secondary Reason for Visit`, "other")) %>%
  mutate(across(starts_with("New or Continued Medication"), ~na_if(., "Not applicable")),
         across(starts_with("Medication #"), ~na_if(., -9))) %>%
  rowwise() %>% 
  mutate("Number of New Medications" = sum(c_across(starts_with("New or Continued Medication")) == "New" & !is.na(c_across(starts_with("New or Continued Medication")))),
         "Number of Continued Medications" = sum(c_across(starts_with("New or Continued Medication")) == "Continued" & !is.na(c_across(starts_with("New or Continued Medication")))))


```

## Assembling the new medication dataset with names

There are over 1,200 medications (generic and non-generic), with unique identifying numbers in the NAMCS data documentation. The code below shows how I replaced the names in an excel spreadsheet I generated from a 20 page pdf, with the codes that were present in the raw NAMCS data.

```{r}
# Create a named vector for the replacements
replacement_vector <- medcodes_clean$`Rx Name`
names(replacement_vector) <- medcodes_clean$`Rx ID Number`

# Replace the values in the medication columns
data_meds_names <- data_meds %>%
  mutate_at(
    vars(starts_with("Medication #")),
    list(~ replacement_vector[.])
  )

# Add the "Total New Medications" and "Total Continued Medications" variables
data_meds_names <- data_meds_names %>%
  rowwise() %>%
  mutate(
    Total_New_Medications = sum(c_across(starts_with("New or Continued Medication")) == "New", na.rm = TRUE),
    Total_Continued_Medications = sum(c_across(starts_with("New or Continued Medication")) == "Continued", na.rm = TRUE)
  ) %>%
  ungroup()

data_meds_names <- data_meds_names %>%
  mutate(Total_New_Medications = as.numeric( Total_New_Medications),
         Total_Continued_Medications = as.numeric(Total_Continued_Medications))

#creating dataset of individuals with hypertension
data_htnyes_mednames <- data_meds_names %>%
  filter(`Major Reason for Visit` %in% c("Chronic problem, routine",
                                         "New problem (less than 3 mos. onset)",
                                         "Chronic problem, flare-up",
                                         "Preventive care")) %>%
  filter(`Hypertension` %in% c("Yes"))

#creating dataset of primary and secondary htn visits with medication names
data_htnvisits_mednames <- data_meds_names %>%
    filter(`Major Reason for Visit` %in% c("Chronic problem, routine",
                                         "New problem (less than 3 mos. onset)",
                                         "Chronic problem, flare-up",
                                         "Preventive care")) %>%
  filter(`Primary Reason for Visit` %in% c("Hypertension", "Blood pressure test") |
           `Secondary Reason for Visit` %in% c("Hypertension", "Blood pressure test"))
```

## Frequencies of Medications across dataset

```{r}
# Reshape the dataset
meds_long_htn <- data_htnyes_mednames %>%
  pivot_longer(cols = starts_with("Medication"),
               names_to = "Medication_Variable",
               values_to = "Medication_Name")%>%
    drop_na(Medication_Name)

#count medications for htn 
medication_counts_htn <- meds_long_htn %>%
  filter(Medication_Name != "At least one medication was reported" & Medication_Name != "No medications were reported") %>%
  mutate(Medication_Name = str_to_title(str_to_lower(Medication_Name))) %>%
  group_by(Medication_Name) %>%
  summarize(Total_Frequency = n(), .groups = 'drop') %>%
  rename(Medication = Medication_Name)


# Reshape the dataset
meds_long_htnvisits <- data_htnyes_mednames %>%
  pivot_longer(cols = starts_with("Medication"),
               names_to = "Medication_Variable",
               values_to = "Medication_Name")%>%
    drop_na(Medication_Name)

#count medications for htn visits
medication_counts_htnvisits <- meds_long_htnvisits %>%
  filter(Medication_Name != "At least one medication was reported" & Medication_Name != "No medications were reported") %>%
  mutate(Medication_Name = str_to_title(str_to_lower(Medication_Name))) %>%
  group_by(Medication_Name) %>%
  summarize(Total_Frequency = n(), .groups = 'drop') %>%
  rename(Medication = Medication_Name)

```



```{r, include=FALSE}
# Arrange the dataset in descending order based on frequency
medication_counts_htn <- medication_counts_htn %>%
  arrange(desc(Total_Frequency))

# Select the top 20 medicines
top_20_medicines_htn <- medication_counts_htn %>%
  slice(1:20)

# Calculate the sum of frequencies for the remaining medicines
other_medicines_frequency_htn <- sum(medication_counts_htn$Total_Frequency[21:nrow(medication_counts_htn)])

# Create a data frame for the "other" category
other_category_htn <- data.frame(Medication = "Other", Total_Frequency = other_medicines_frequency_htn)

# Combine the top 20 medicines with the "other" category
top20_medication_counts_htn <- rbind(top_20_medicines_htn, other_category_htn)

# View the final dataset
print(top20_medication_counts_htn)

total_sum_htn <- sum(top20_medication_counts_htn$Total_Frequency[1:20])

# Add a new column with the percentage for each count
top20_medication_counts_htn$Percentage <- (top20_medication_counts_htn$Total_Frequency / total_sum_htn) * 100

# Order Medication by Percentage and create a factor with ordered levels
top20_medication_counts_htn$Ordered_Medication <- factor(top20_medication_counts_htn$Medication,
                                                      levels = top20_medication_counts_htn$Medication[order(-top20_medication_counts_htn$Percentage)])

my_colors <- c("#5DA5DA", "#FAA43A", "#60BD68", "#F17CB0", "#B2912F", 
               "#B276B2", "#DECF3F", "#F15854", "#4D4D4D", "#9F9F9F",
               "#B8E186", "#FFAFAF", "#AEC7E8", "#FFB347", "#D7B5DC",
               "#C7C7C7", "#E377C2", "#7F7F7F", "#66C2A5", "#FC8D62")

bar_graph <-ggplot(top20_medication_counts_htn[1:20,], aes(x = Ordered_Medication, y = Total_Frequency, fill = Ordered_Medication)) +
  geom_bar(width = 0.7, stat = "identity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Top 20 Medication Counts",
       x = "Medication",
       y = "Total Frequency",
       fill = "Medication") +
  geom_text(aes(label=Total_Frequency), vjust=-0.3, size=3.5) +
  scale_fill_manual(values = my_colors) +
  guides(fill = guide_legend(title = NULL)) +
  theme(panel.grid.major.y = element_line(color = "grey", linetype = "dashed"))



```

## Most frequent medicines used by people with hypertension

```{r}
#| echo: false
print(bar_graph)

```

## Table 1 for those who have primary or secondary reason for visit that is related to htn

```{r entry 5}

# Continuous and categorical variables
cont_vars <- c("Age", "Number Medications", "Time with Doctor", "Systolic Blood Pressure")
cat_vars <- c("Race and Ethnicity", "Smoking History", "Major Reason for Visit", "Primary Care",
              "Hypertension", "Chronic Kidney Disease", "Congestive Heart Failure", 
              "Diabetes Type 1", "Diabetes Type 2", "Obesity", "Medication", 
              "Recommended for ED or ER")

# Create the summary table
my_table <- CreateTableOne(vars = c(cont_vars, cat_vars), data = data_htnvisits, factorVars = cat_vars)

# Print the table
print(my_table, smd = TRUE)

```

## A few descriptive plots

```{r entry 6}

#age distribution by race and ethnicity
ggplot(data_htnvisits, aes(x = `Race and Ethnicity`, y = Age, fill = `Race and Ethnicity`)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Age Distribution by Race and Ethnicity",
       x = "Race and Ethnicity",
       y = "Age")

ggplot(data_htnvisits, aes(x = `Smoking History`, fill = `Smoking History`)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Frequency of Smoking History",
       x = "Smoking History",
       y = "Count")

ggplot(data_htnvisits, aes(x = `Number Medications`, y = `Systolic Blood Pressure`)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  theme_minimal() +
  labs(title = "Systolic Blood Pressure vs Number of Medications",
       x = "Number of Medications",
 get      y = "Systolic Blood Pressure")


data_htnvisits %>%
  ggplot(aes(x = `Race and Ethnicity`, y = `Number Medications`, fill = `Race and Ethnicity`)) +
  geom_boxplot() +
  labs(title = "Number of Medication by Race and Ethnicity")

```

## Modeling Data for those who had a primary/secondary visit reason as blood pressure test or HTN

```{r entry 8}
#Split the data into training and testing sets
set.seed(123)
data_split <- initial_split(data_htnvisits, prop = 0.75)
train_data <- training(data_split)
test_data <- testing(data_split)

#Create a recipe, had to add step nzv to remove variables with near-zero variance
htnvisits_recipe <- recipe(`Number Medications` ~ ., data = train_data) %>%
  step_nzv(all_predictors()) %>%
  step_normalize(all_numeric(), -all_outcomes()) %>%
  step_dummy(all_nominal(), -all_outcomes())


#Choose and set up a model
linear_model <- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression")

#Create a workflow
htnvisits_workflow <- workflow() %>%
  add_recipe(htnvisits_recipe) %>%
  add_model(linear_model)

#cross validation
set.seed(123)
cv_results <- htnvisits_workflow %>%
  fit_resamples(
    resamples = bootstraps(train_data, times = 10),
    metrics = metric_set(rmse, rsq)
  )
cv_results %>%
  collect_metrics()

#Visualize the performance of the model during cross-validation.
cv_results %>%
  collect_metrics() %>%
  ggplot(aes(x = .metric, y = mean, ymin = mean - std_err, ymax = mean + std_err)) +
  geom_pointrange() +
  labs(title = "Cross-validation Metrics (with Standard Error)",
       x = "Metric",
       y = "Value") +
  theme_minimal()

#train the final model
final_model <- htnvisits_workflow %>%
  fit(data = train_data)

#evaluate the model
test_results <- final_model %>%
  predict(test_data) %>%
  bind_cols(test_data) %>%
  metrics(truth = `Number Medications`, estimate = .pred)

#Visualize the model's predicted values against the actual values for the test set.
test_predictions <- final_model %>%
  predict(test_data) %>%
  bind_cols(test_data)

ggplot(test_predictions, aes(x = `Number Medications`, y = .pred)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Actual vs. Predicted Number Medications",
       x = "Actual Number Medications",
       y = "Predicted Number Medications") +
  theme_minimal()

#interpret the model
final_model %>%
  pull_workflow_fit() %>%
  tidy()

```

## Looking at the Prescriptions used for Dashboards

```{r entry 7, eval=FALSE, include=FALSE}

make a data set called "new medications" using the raw file
include the variables "MED1, MED2, MED3...through MED30" do the same for "NCMED1, NCMED2... through NCMED30"
```


